Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-08-26 draft.tex
Caption of centerline figure done

\subsection{Image Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/centerline}
  \caption{Centerline}
  \label{fig:centerline}
\end{figure}
\com{Image Data}
Four trained technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model (FEM) is initialized using three user-placed land marks.
%%
A click on the L3 vertebra center gives the model the initial position, two clicks on the top and bottom of the vertebra describes an initial model height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline} (a)).
%%
Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline} (b)):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from the course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

>>>>

\subsection{Image Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/centerline}
  \caption{
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	%%
	(a) right: FEM aligned to a MRI data set.
	%%
	The purple dashed line represents the centerline, describing the lumbar spine canal with 93 points.
	%%
	(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
	}
  \label{fig:centerline}
\end{figure}
\com{Image Data}
Four trained technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-placed land marks.
%%
A click on the L3 vertebra center gives the model the initial position, two clicks on the top and bottom of the vertebra describes an initial model height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
%%
Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline}~b):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from the course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.
